Analyzing Trends in U.S. Residential Real Estate Prices

### 2024-04-17 outputs/paper/paper.qmd
paper final draft

Data used in this paper was cleaned, processed and tested with the programming language R [@citeR]. Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate], `maps` [@citeMaps], `mgcv` [@citeMgcv], `Arrow` [@citeArrow].

@@

Our study focuses on U.S. house price trends using the extensive datasets available from Zillow Research [@citeZillowdata]. This data repository provides a broad array of real estate statistics, capturing the dynamics of house prices across various regions in the United States. As a cornerstone of transparency and public engagement, Zillow Research offers open access to these vital real estate data, which are instrumental in understanding housing market behaviors.

The dataset encompasses detailed information on house prices over several years, including breakdowns by state, city, and occasionally neighborhood levels, from median list prices to the Zillow Home Value Index (ZHVI). This level of detail facilitates a nuanced analysis of house price trends and the identification of significant market dynamics, which can guide potential real estate investments and policy formulations.

@@

@@

Our analysis began with the critical task of refining the raw dataset downloaded from Zillow. The initial phase involved understanding the dataset’s structure. As shown in @tbl-raw_data, this dataset typically includes multiple attributes such as Region ID, Region Type, location (states), year of transaction, house price, etc. Each record represents a unique property, providing a detailed snapshot of its characteristics.

@@

$$\text{Price} = \beta_0 + \beta_1 \times \text{Year} + \beta_2 \times \text{NumBedroom}$$

@@

$$\text{Price} = \beta_0 + \beta_1 \times \text{Year} + \beta_2 \times \text{Year}^2 + \beta_3 \times \text{NumBedroom} + \beta_4 \times \text{NumBedroom}^2$$

@@

$$\text{Price} = -13,969,628.019 + 6,958.769 \times \text{Year} + 59,897.327 \times \text{NumBedroom}$$

@@

$$\text{Price} = 212,976.031 + 3,866,485.706 \times \text{Year} + 1,601,816.793 \times \text{Year}^2$$
$$+ 6,591,537.312 \times \text{NumBedroom} + 1,365,656.076 \times \text{NumBedroom}^2$$

@@

The escalation in prices may coincide with inflationary trends affecting various economic sectors, suggesting that housing markets are susceptible to the impacts of global monetary policies and financial climates. These inflationary pressures can distort consumer purchasing power, creating a complex scenario where rising prices may not necessarily reflect actual increases in property value. The data also suggests that demographic shifts, particularly changes in family structures with more members per household, significantly influence demand and, consequently, prices in the housing market.

@@

The increase in demand for larger homes, as indicated by the data, suggests a potential shift in housing preferences, possibly spurred by the recent pandemic, which emphasized the need for multifunctional living spaces that combine comfort with functionality. This shift reflects not only a preference for larger living spaces but also an adaptation to new realities—homes that can accommodate remote work, virtual schooling, and indoor recreation.

@@

The insights derived from the geographical distribution of housing markets highlight opportunities for strategic infrastructural investments. Identifying areas with rapid price growth allows stakeholders to pinpoint locations primed for development or in need of policy adjustments to stabilize the market. Additionally, this data facilitates the targeted allocation of resources to underserved regions, promoting a more balanced economic distribution and enhancing overall economic equity.

>>>>

Data used in this paper was cleaned, processed, modeled and tested with the programming language R [@citeR]. Also with support of additional packages in R: `tidyverse` [@citeTidyverse], `ggplot2` [@citeGgplot], `janitor` [@citeJanitor], `readr` [@citeReadr], `knitr` [@citeKnitr], `modelsummary` [@citeModelsummary], `testthat` [@citeTestthat], `KableExtra` [@citeKableEx], `viridis` [@citeViridis], `lubridate` [@citeLubridate], `maps` [@citeMaps], `mgcv` [@citeMgcv], `Arrow` [@citeArrow].

@@

Our study focuses on U.S. house price trends using the extensive datasets available from Zillow Research [@citeZillowdata]. This data repository provides a broad array of real estate statistics, capturing the dynamics of house prices across various regions in the United States. The dataset encompasses detailed information on house prices from 2000 to 2024, including breakdowns by state, city, and occasionally neighborhood levels, from median list prices to the Zillow Home Value Index (ZHVI). As a cornerstone of transparency and public engagement, Zillow Research offers open access to these vital real estate data, which are instrumental in understanding housing market behaviors.

@@

## Measurement

In this study, the conversion of real-world real estate transactions into quantifiable data entries is meticulously executed using Zillow's extensive datasets. Each real-world transaction is represented as a data point in our analysis, capturing essential attributes such as sale price, property location, and characteristics like the number of bedrooms and bathrooms. For example, a house sale in Miami with three bedrooms and two bathrooms, sold at a specific price, is recorded with precise values for each attribute, including the sale date. This method ensures that each entry in our dataset accurately reflects an actual event, maintaining the integrity and accuracy of the data. The rigorous methodology employed in documenting these transactions allows for precise, data-driven insights into market dynamics, providing a solid foundation for analyzing trends and drawing conclusions about factors influencing real estate prices.

@@

Our analysis began with the essential task of refining the raw dataset downloaded from Zillow. The original data was largely retained because it covers a time span that aligns well with the needs of this study. The initial phase involved understanding the dataset’s structure. As demonstrated in @tbl-raw_data, this dataset typically includes multiple attributes such as Region ID, Region Type, location (states), year of transaction, house price, etc. Each record represents a unique property, offering a detailed snapshot of its characteristics.

@@

$$\text{Price} = \beta_0 + \beta_1 \times \text{Year} + \beta_2 \times \text{NumBedroom} + \varepsilon$$

@@

$$\text{Price} = \beta_0 + \beta_1 \times \text{Year} + \beta_2 \times \text{Year}^2 + \beta_3 \times \text{NumBedroom} + \beta_4 \times \text{NumBedroom}^2 + \varepsilon$$

@@

$$\hat{\text{Price}}  = -13,969,628.019 + 6,958.769 \times \text{Year} + 59,897.327 \times \text{NumBedroom}$$

@@

$$\hat{\text{Price}}  = 212,976.031 + 3,866,485.706 \times \text{Year} + 1,601,816.793 \times \text{Year}^2$$
$$+ 6,591,537.312 \times \text{NumBedroom} + 1,365,656.076 \times \text{NumBedroom}^2$$

@@

The modeling efforts in this paper underscore the complexity of the housing market's response to economic forces and suggest that straightforward economic models might not always suffice in capturing the real-world dynamics of real estate pricing. The escalation in prices may coincide with inflationary trends affecting various economic sectors, suggesting that housing markets are susceptible to the impacts of global monetary policies and financial climates. These inflationary pressures can distort consumer purchasing power, creating a complex scenario where rising prices may not necessarily reflect actual increases in property value. The data also suggests that demographic shifts, particularly changes in family structures with more members per household, significantly influence demand and, consequently, prices in the housing market.

@@

The increase in demand for larger homes, as indicated in previous sections, suggests a potential shift in housing preferences, possibly spurred by the recent pandemic, which emphasized the need for multifunctional living spaces that combine comfort with functionality. This shift reflects not only a preference for larger living spaces but also an adaptation to new realities—homes that can accommodate remote work, virtual schooling, and indoor recreation.

@@

The insights derived from the geographical distribution of housing markets underscore opportunities for strategic infrastructural investments. The correct utilization of the Generalized Additive Model (GAM) was instrumental in identifying subtle patterns, such as the impact of microeconomic factors like local zoning laws or the availability of amenities on housing prices. Identifying areas with rapid price growth enables stakeholders to pinpoint locations primed for development or in need of policy adjustments to stabilize the market. Additionally, this data facilitates the targeted allocation of resources to underserved regions, promoting a more balanced economic distribution and enhancing overall economic equity.
